Complex and informative dependencies propagate throughout ecology and can be revealed using copulas

### 2019-05-13 Paper.Rmd
more edits

The importance of a complete copula description of dependence between variables for ecology and related fields

The importance of a complete description of dependence between variables for ecology and related fields

Complex, informative, and previously ignored dependencies propogate throughout ecology and are revealed using copulas

Complex, informative, and previously ignored dependencies propagate throughout ecology and can be revealed using copulas

The complexity of ecological relationships is revealed by copulas

The true complexity of ecological relationships is revealed by copulas

The true complexity of ecological relationships is only revealed by copulas

Copulas reveal complex, informative, and previously ignored dependence
structures between variables across ecological disciplines

Copulas reveal complex and informative dependence
structures between variables across ecological disciplines

Copulas reveal complex, informative, and previously ignored dependence
structures which in turn illuminate multiple ecological disciplines

Copulas reveal complex, informative, and previously ignored dependence
structures which in turn can illuminate every subdiscipline of ecology

Copulas reveal complex, informative, and previously ignored dependence
structures which in turn can illuminate every subfield of ecology

Copulas reveal complex, informative and previously ignored dependence
structure between variables throughout ecology

The complexity of dependencies between ecological variables, previously
ignored, is revealed by copulas and in turn illuminates multiple 
disciplines of ecology

\noindent \emph{Keywords:}

\noindent dependence, regression, correlation, copula, population, community

\noindent \emph{Type of Article:}

\noindent Ideas and Perspectives

@@

\noindent

@@

\noindent \emph{The number of words in each text box:}

\noindent No text boxes

@@

\noindent \emph{The number of figures, tables, and text boxes:}

\noindent X figures, X tables, 0 text boxes

\noindent \emph{The name and complete mailing address (including telephone and fax numbers and e-mail address) of the person to whom correspondence should be sent:}

\noindent Daniel Reuman, 2101 Constant Ave, Lawrence, KS, 66047, reuman@ku.edu, 626 560 7084.

@@

All branches of ecology 
study relationships among environmental and biological 
variables. However, commonly used correlation and regression 
approaches to studying 
such relationships are limited - standard 
approaches provide only a very partial description of a relationship between variables. 
For instance, datasets showing markedly different 
relationships may nevertheless have the same correlation coefficient value 
(Fig. \ref{fig_cop_pedag1}A-B). The variables of Fig. \ref{fig_cop_pedag1}A 
(respectively, Fig. \ref{fig_cop_pedag1}B) are principally 
related in the left (respectively, right) portions of their distributions, 
an asymmetric pattern of dependence that can
have profound ecological significance, as discussed below, but is not captured by 
correlation. 
Monotonic transformations can alter the Pearson, but not the Spearman
correlation (Appendix \ref{SM-pedagog1_details}, 
Fig. \ref{SM-fig_transf_pearson_spearman}). Although correlation
is only one way to 
study relationships between variables, other common metrics also 
provide partial information.

\begin{figure}[!h]
\begin{center}
\includegraphics[width=3.5in]{./Results/PedagogFig1.pdf}
\caption{(A-D) Bivariate datasets showing diverse relationships between 
variables; many of these datasets nevertheless have the same Pearson (P)
or Spearman (S) correlation coefficients. (C, D) Contours 
of density functions of Clayton (C) and survival-Clayton 
(D) copulas with parameter 5 (see section on methods for the observational hypothesis), together
with samples from the corresponding distributions. (E,F) Reflections
of C, D, respectively, about a vertical axis. (G) Contours of the
density function of a normal copula (methods for the observational hypothesis) with 
parameter chosen to
have the same Spearman correlation as A-D, together with a sample.
(H) The portion of the Spearman correlation within a band (diagonal lines
show two bands, the data in the left/lower band showing stronger
dependence than the right/upper band) can be computed 
for any band (methods for the observational hypothesis) to describe how dependence
varies from one part of the two distributions to another.
Subscripts h and v 
stand for "horizontal" and "vertical". \label{fig_cop_pedag1}}
\end{center}
\end{figure}

Well-developed approaches do exist, however, and are 
applied widely in finance, neuroscience, bioinformatics and other fields, 
that provide a description of the dependence 
between two or more variables that can be mathematically proven 
to be complete [@joe2014_dependence; @nelsen2006_copula]. 
The approaches are based on 
the statistical concept of the *copula*, which we here 
introduce. Copula approaches separate the information contained in a bivariate 
random variable into two distinct parts: the information contained in the 
marginal distributions (which says nothing about the dependence between the 
variables), and the remaining information (which is solely about
the dependence). From a data perspective, given 
a sample $(x_i,y_i)$, $i=1,2,\ldots,n$ (e.g., 
Fig. \ref{fig_cop_pedag1}A-B), the associated dependence
structure is revealed by the bivariate plot of $u_i$ versus $v_i$, where
$u_i$ is the rank of $x_i$ in the set of the $x_j$ ($j=1,\ldots,n$), divided by
$n+1$; and $v_i$ is defined in the same way but using the $y_i$. The $u_i$ and 
$v_i$ are called *normalized ranks* of the $x_i$ and $y_i$. See, 
for instance,
Fig. \ref{fig_cop_pedag1}C-D (the points), which show the copula structures of 
A-B, respectively. Ranking makes the marginal distributions of 
the underlying random variables uniform, isolating the 
dependence structure only.
Probability density 
functions (pdfs) corresponding to the copulas associated 
with Fig. \ref{fig_cop_pedag1}C-D are superimposed on those panels. 
See @nelsen2006_copula and @joe2014_dependence
for general background. A theorem of @sklar1959_theorem
states that copulas give a unique and complete 
representation of the dependence
between variables.

A principal benefit of a copula approach is 
that it can sensitively detect tail dependence and asymmetries of tail dependence. 
Tail dependence is association between extreme values of variables. 
If smaller values of two positively associated variables are more 
strongly associated than are larger values, the variables are said to
have stronger lower- or left-tail dependence than right- or upper-tail dependence
(Fig. \ref{fig_cop_pedag1}A, C);
and \emph{vice versa} if larger values are more strongly associated than smaller
ones (Fig. \ref{fig_cop_pedag1}B, D). Datasets of the same 
correlation can have a range of tail dependence 
characteristics (Fig. \ref{SM-fig_cop_pedag}), which can be quantified (see section on
methods for the observational hypothesis).

The goal of this paper is to illustrate the potential for 
broader application of copulas in ecology. 
Our results provide initial support for the following 
overall hypothesis: environmental, ecological and evolutionary processes 
generate and transmit copula structure and tail dependence, with substantial 
consequences for theoretical and conceptual understanding, for inference, and for 
applications across ecology. This is a broad hypothesis, and 
even the substantial empirical and theoretical
evidence we present here has only begun to test it. So we hope our work stimulates further
evaluation of the hypothesis by the research community.
Our overall hypothesis is examined by assessing several more specific 
hypotheses, described below and outlined in Fig. 
\ref{MasterFigure}, all of which bear upon the overall hypothesis.

\begin{figure}[!h]
\begin{center}
\includegraphics[width=7in]{./MasterFigure.pdf}
\caption{Guide for hypotheses addressed in the text and how each instantiates the
overall hypothesis. Labels are defined 
in the text to refer to parts of the figure. "Causal hypotheses" (i.e., hypotheses about 
causes of copula structure) are labeled with letters A-D. "Consequential 
hypothesis" (i.e., hypothesis about 
consequences of copula structure) are labeled with letters X-Z. The "observational 
hypothesis" (i.e., our hypothesis about 
copula structure of data) is labeled with O.\label{MasterFigure}}
\end{center}
\end{figure}

@@

Because the observational hypothesis bears upon all the causal
and consequential hypotheses, we test it first, in sections 2-4 of the paper.
We analyze several datasets, including 
environmental, species-trait, phenological, population, community, and 
ecosystem functioning data, selected to span multiple 
sub-fields and organizational levels of ecology. 
Sections 5 and 6 test the causal hypotheses, and sections 7 and 8 test the
consequential hypotheses. Section 9 is the Discussion.
Copula approaches have been used to great effect in neuroscience [@onken2009], 
bioinformatics [@Kim2008], medical research [@Emura2016], direct study of environmental 
variables [@Serinaldi2008; @She2018; @Goswami2018; @li2013], and finance [@Li2000]. 
To our knowledge, these concepts have never been used in ecology. 
We argue that benefits to ecology of using a complete 
copula desription of dependence will be substantial.

@@

dt <- tibble(Description = c("1. p-value, independence test",
           "2. Minimum-AIC copula",
           "3. AIC weight of minimum-AIC copula",
           "4. AIC weight of N copula",
           "5. p-value, Cramer-von Mises goodness of fit test",
           "6. p-value, Kolmogorov-Smirnov goodness of fit test",
           "7. Model averaged lower-tail dependence",
           "8. Model averaged upper-tail dependence",
           "9. Model averaged lower- minus upper-tail dependence",
           "10. Asymmetry rank in surrogates")
)
tab_bivar_summary<-cbind(dt,tab_bivar_summary)
knitr::kable(tab_bivar_summary[1:9,], 
             format="latex",
             caption = "Summary of OH results for bivariate datasets. The $p$-values in rows 5-6 are for the minimum-AIC copula. Model averaging used for rows 7-9 was based on AIC weights. Cedar Creek results are for the year 2000.\\label{tab_bivar_summary}",
             booktabs=T, linesep = "\\addlinespace",
             col.names = c(" ", "Soil C and N", "Bird masses and BMR", "Mammal masses and BMR", "Cedar Creek data")
             )%>%column_spec(2:5,width="1.4cm")%>%
             column_spec(1, width = "8.5cm")%>%kable_styling(latex_options = "basic",position = "center")
```

@@

For the model we examined, causal hypothesis D (Fig. \ref{MasterFigure}) was 
correct that asymmetric tail dependence in evolutionary changes 
can produce similarly asymmetric tail dependence in character distributions across
phylogeny tips (Appendix \ref{SM-evol_mech}). This topic is revisited in the Discussion.

@@

<!--what we did and why is it important to study?-->
We showed that non-N copula structure and asymmetric tail dependence are 
common in datasets across multiple sub-disciplines
and levels of organization in ecology (Fig. \ref{MasterFigure}, O), although no published 
research we are aware of has 
taken this structure into account. We hypothesized four 
mechanisms (Fig. \ref{MasterFigure}, A-D) that may cause non-N copula structures. One
mechanism (A) was via nonlinear environmental influences
on ecological variables. A second mechanism (B) produces tail dependence via
a Moran effect, whereby dependencies between environmental variables 
beget similar dependencies between ecological variables in different locations. 
We discuss below how commonly these and other
potential mechanisms may opperate. 
We also hypothesized and demonstrated important consequences of 
non-N and tail-dependent copula structure in ecology 
(Fig. \ref{MasterFigure}, X-Z). First (X), the skewness
of a spatial-average time series is influenced by tail dependencies between its 
constituent time series. Thus right-tail-dependent local time series 
lead to "spiky" spatially averaged time series, with large outbreaks.
These observations could have implications for numerous exploited and pest species,
for which the spatial average time series have applied importance.
Second (Y), metapopulation extinction risk can be accentuated or reduced by tail dependence
properties of environmental variables. We also showed effects of copula 
structure on Taylor's law (Z), which has been applied in many fields.
In our view, our results make our "overall hypothesis," framed in the Introduction, quite 
reasonable and worthy of broader testing by the research community. We have illustrated 
extensively how such tests might be carried out by performing them ourselves on 
example datasets and models. Further suggestions future possible tests are below. 
<!--***DAN: make sure you have enough of this, Dan-->
It already seems likely that comprehensive understanding 
of many ecological phenomena is impossible without 
a complete copula characterization of dependence between variables.

@@

<!--a para on how likely one is to get copula structure via the Liebig mechanism-->
Our first causal mechanism (Fig. \ref{MasterFigure}, A) seems likely to operate commonly.
Liebig's law and the idea of a "limiting 
nutrient" of growth are dominant paradigms in ecology.
Although we did not systematically test the first two parts of causal 
hypothesis A, many studies document nonlinear influences of environmental
variables on ecological quantities at all levels of organization. 
These parts of A could be tested
straightforwardly for numerous datasets by examing the copulas of environmental and ecological
variables (see below discussion about the green spruce aphid, and 
Appendix \ref{SM-sect:greenspruce}, which illstrate how to do this).

<!--***DAN: stoped here and went back to put in Shya/Lawrence recent results with 
aphid counts and winter temp-->

<!--para on how likely one is to get copula structure via the Moran mechanism-->
There are also reasons to expect that our Moran mechanism (Fig. \ref{MasterFigure}, B) for 
copula structure and tail dependence will
operate commonly, because Moran effects are common [@Liebhold2004], and 
non-N and tail-dependent copula structures are often found in 
environmental variables. 
If large meteorological events are 
also spatially widespread, then
one would expect environmental variables associated with these events to take extreme
values simultaneously across large spatial areas, producing tail dependence in measurements
made through time at different locations. Non-extreme values, not associated with extreme, 
spatially extensive weather events, may
be more associated with local phenomena, and therefore 
less prone to being correlated across large areas.
For instance, @Serinaldi2008 examined the spatial dependence 
of rainfall observations in Central Italy. G or Student 2-copulas 
were possible candidates for modelling dependence, and 
neither of these is an N copula.
A long-term study (1950–2014) in the Loess Plateau of China [@She2018]
showed that a G copula (which has upper tail dependence) could effectively model 
the spatial dependence of drought variables in the study area.
Bivariate copula analysis was also used in forecasting the 
co-occurrence of extreme events (flood or drought) 
over the North Sikkim Himalayas, India, using spatial datasets [@Goswami2018].

<!--Holder stuff-->
Our simulations of character evolution suggest the hypothesis that evolutionary
changes through time in bird and mammal BMR and body size may exhibit right-tail dependence,
contrary to standard assumptions of models of character evolution. This is
a hypothesis only, because the right-tail dependence of Fig. \ref{fig_biv_multi_raw_cop}F,G
could have come about in another, unknown way: our simulations show that tail dependence 
in evolutionary changes is sufficient, but may not be necessary, to produce tail
dependence in characters of extant species. For instance, as discussed below and in
Appendix \ref{SM-sect:missingdata} (also Fig. \ref{SM-fig:missingdata}), 
systematically missing data can produce tail dependence and may
have influenced results for the BMR-body mass datasets. Even if the hypothesized 
evolutionary mechanism
is correct, our results simply replace one question, i.e., why do we see right-tail
dependence between BMR and body mass for birds and mammals, with another question,
i.e., why might we see right-tail dependence in evolutionary changes in these traits?
See Appendix \ref{SM-evol_mech} for additional thoughts on the potential importance of 
copulas for studies of character evolution.

<!--Discuss stuff here on tradeoff and life history that Bever told me about-->
The relationships we studied between BMR and body mass relate
to a trade-off between mass-specific BMR (i.e., BMR per unit body mass) and body
mass itself. Trade-offs between life history characteristics are a focal topic in 
ecology. Copulas probably interrelate with 
life-history theory in additional ways beyond what we could demonstrate 
with the data we used.
It is well known that energy allocation to a life function, X (e.g., reproduction)
will reduce the energy that can be allocated to other functions, $Y_1$, $Y_2$, $Y_3$ 
(e.g., growth, predation avoidance; this is called the principle of 
allocation).
But it is equally well known that X can trade-off against any or all of 
the $Y_i$. Therefore, for large values of X which approach absolute 
limitations, there may be a strong 
association between X and $Y_1$. Whereas for small values of X there may be
little association - resources not allocated to X can instead be allocated 
to $Y_1$, $Y_2$, $Y_3$ or to some combination. This constitutes 
asymmetric tail dependence between X and $Y_1$. For instance, @Winemiller1992
described a three-way trade-off in North American fishes between age of reproductive
maturity, juvenile survivorship, and fecundity. The trade-off should, in theory,
produce a tight association between age of maturity and fecundity for 
fishes with low age of maturity, but little association between those
variables for later-maturing
fishes because those fishes may invest much or little in juvenile survival.
Although copulas may be very useful for studying multi-dimensional trade-offs, 
applications will 
require careful attention to the possible consequences of biased sampling:
if the degree of completeness of a dataset is associated with one or
more of the characters being studied then statistical artefacts
may bias conclusions (Appendix \ref{SM-sect:missingdata}, 
Fig. \ref{SM-fig:missingdata}).

@@

<!--***DAN: green spruce aphid all the way from causes to consequences-->
Our hypotheses cover propagation of copula structure from causes, 
to observable data, to consequences for human concerns.
We here take a closer look at the green spruce aphid, which illustrates 
all these stages in one example.
Green spruce aphid abundance, as detected in the suction trap data we use (Data,
Appendix \ref{SM-Data_details}),
is known to be strongly influenced by temperature the previous winter (@Sheppard2016,
supplementary fig. 6).
For each of our 10 sampling locations, we examined (Appendix \ref{SM-sect:greenspruce}, 
Fig. \ref{}) the copula of winter temperature
and green spruce aphid abundance time series for the location, finding 
substantial left-tail dependence: apparently winter temperature has an asymmetric
influence on abundance in the sense that cold winters produce low abundances
but warm winters do not yield especially higher abundances than moderate winters
(in passing, this is evidence for the first two parts of causal mechanism A).

<!--***DAN: could remove the next para if you need to save words-->
<!--a para on stability of the total biomass time series in a community, and skewness - sets up Shya's later paper as future work-->
Our results showed that the skewness of the 
spatial average of local time series from different locations is influenced 
by their tail dependence. But the logic of this result should also apply to any collection of time
series, whether associated with different locations in space or not, that are summed or averaged. 
Another potential application is time series of all species from a single community, e.g., all
plants in a quadrat surveyed repeatedly over time. A huge 
literature has focused on synchrony versus 
compensatory dynamics between such time series, and 
the influence of such interspecific relationships on the 
variability, through time, of total-community
properties such as total biomass. Typically, variability through time of total-community
properties is measured with the coefficient of variation, but the skewness of such a time series
may also be of interest because it can characterize the extent to
which community properties show "spikiness". Future work on copula structure of interspecific relationships in communities and its implications for community stability seems likely to be an exciting avenue of research.

>>>>

Copulas and their importance for ecology

Copulas and their importance throughout ecology

Copulas and their potential for ecology

Copulas and their potential throughout ecology

Copulas and their importance

Copulas in ecology

Copulas

\noindent \emph{Keywords:} dependence, regression, correlation, copula, population, community

\noindent \emph{Type of Article:} Ideas and Perspectives

@@

@@

\noindent \emph{The number of words in each text box:} No text boxes

@@

\noindent \emph{The number of figures, tables, and text boxes:} 7 figures, 3 tables, 0 text boxes

\noindent \emph{The name and complete mailing address (including telephone and fax numbers and e-mail address) of the person to whom correspondence should be sent:} Daniel Reuman, 2101 Constant Ave, Lawrence, KS, 66047, reuman@ku.edu, 626 560 7084.

@@

All branches of ecology 
study relationships among environmental and biological 
variables. However, commonly used correlation and regression 
approaches to studying 
such relationships are limited, and 
provide only a partial description of a relationship between variables. 
For instance, datasets showing markedly different 
relationships may have the same correlation coefficient
(Fig. \ref{fig_cop_pedag1}A-B). The variables of Fig. \ref{fig_cop_pedag1}A 
(respectively, Fig. \ref{fig_cop_pedag1}B) are principally 
related in the left (respectively, right) portions of their distributions, 
an asymmetric pattern of dependence that can
have profound significance, as discussed below, but is not captured by 
correlation. 
Monotonic transformations can alter the Pearson, but not the Spearman
correlation (Appendix \ref{SM-pedagog1_details}, 
Fig. \ref{SM-fig_transf_pearson_spearman}). Although correlation
is only one way to 
study relationships between variables, other common metrics also 
provide partial information.

\begin{figure}[!h]
\begin{center}
\includegraphics[width=3.5in]{./Results/PedagogFig1.pdf}
\caption{(A-D) Bivariate datasets showing diverse relationships between 
variables; many of these datasets nevertheless have the same Pearson (P)
or Spearman (S) correlation coefficients. (C, D) Contours 
of probability density functions (pdf) of Clayton (C) and survival-Clayton 
(D) copulas with parameter 5 (see section on methods for the observational hypothesis), together
with samples from the corresponding distributions. (E,F) Reflections
of C, D, respectively, about a vertical axis. (G) Contours of the
pdf function of a normal copula (methods for the observational hypothesis) with 
parameter chosen to
have the same Spearman correlation as A-D, together with a sample.
(H) The portion of the Spearman correlation within a band (diagonal lines
show two bands, the data in the left/lower band showing stronger
dependence than the right/upper band) can be computed 
for any band (methods for the observational hypothesis) to describe how dependence
varies from one part of the two distributions to another.
Subscripts h and v 
stand for "horizontal" and "vertical". \label{fig_cop_pedag1}}
\end{center}
\end{figure}

Well-developed approaches do exist, however, and are 
applied widely in other fields, 
that provide a description of the dependence 
between two or more variables that can be mathematically proven 
to be complete [@joe2014_dependence; @nelsen2006_copula]. 
The approaches are based on 
the concept of the *copula*, which we here 
introduce. Copula approaches separate the information of a bivariate 
random variable into two distinct parts: the information in the 
marginal distributions (which says nothing about the dependence between the 
variables), and the remaining information (which is solely about
the dependence). From a data perspective, given 
a sample $(x_i,y_i)$, $i=1,2,\ldots,n$ (e.g., 
Fig. \ref{fig_cop_pedag1}A-B), the dependence
structure is revealed by the bivariate plot of $u_i$ versus $v_i$, where
$u_i$ is the rank of $x_i$ in the set $x_j$ ($j=1,\ldots,n$), divided by
$n+1$; and $v_i$ is defined in the same way but using the $y_i$. The $u_i$ and 
$v_i$ are called *normalized ranks* of the $x_i$ and $y_i$. See, 
for instance,
Fig. \ref{fig_cop_pedag1}C-D (the points), which show the copula structures of 
A-B, respectively. Ranking makes the marginal distributions of 
the underlying random variables uniform, isolating the 
dependence structure.
See @nelsen2006_copula and @joe2014_dependence
for general background. A theorem of @sklar1959_theorem
states that copulas give a unique and complete 
representation of the dependence
between variables.

A main benefit of a copula approach is 
that it can sensitively detect tail dependence and asymmetries of tail dependence. 
Tail dependence is association between extreme values of variables. 
If smaller values of two positively associated variables are more 
strongly associated than are larger values, the variables are said to
have stronger lower- or left-tail dependence than right- or upper-tail dependence
(Fig. \ref{fig_cop_pedag1}A, C);
and \emph{vice versa} if larger values are more strongly associated than smaller
ones (Fig. \ref{fig_cop_pedag1}B, D). Datasets of the same 
correlation can have a range of tail dependence 
(Fig. \ref{SM-fig_cop_pedag}), which can be quantified (see section on
methods for the observational hypothesis).

The goal of this paper is to illustrate the potential for 
applications of copulas in ecology. 
Our results provide initial support for the following 
overall hypothesis: environmental, ecological and evolutionary processes 
generate and transmit copula structure and tail dependence, with substantial 
consequences for theoretical and conceptual understanding, for inference, and for 
applications across ecology. This is a broad hypothesis, and 
even the substantial empirical and theoretical
evidence we present here has only begun to test it. So we hope our work stimulates further
evaluation of the hypothesis by the research community.
We examine our overall hypothesis by assessing several more specific 
hypotheses, described below and outlined in Fig. 
\ref{MasterFigure}, all of which bear upon the overall hypothesis.

\begin{figure}[!h]
\begin{center}
\includegraphics[width=7in]{./MasterFigure.pdf}
\caption{Guide for hypotheses addressed in the text and how each instantiates the
overall hypothesis. Labels are defined 
in the text to refer to parts of the figure. "Causal hypotheses" (i.e., hypotheses about 
causes of copula structure) are labeled with letters A-D. "Consequential 
hypothesis" (i.e., hypothesis about 
consequences of copula structure) are labeled with X-Z. The "observational 
hypothesis" (i.e., our hypothesis about 
copula structure of data) is labeled O.\label{MasterFigure}}
\end{center}
\end{figure}

@@

Because the observational hypothesis bears upon all the causal
and consequential hypotheses, we test it first, in sections 2-4 of the paper.
We analyze several datasets, including 
environmental, species-trait, phenological, population, community, and 
ecosystem functioning data, selected to span multiple 
sub-fields and organizational levels of ecology. 
Sections 5 and 6 test the causal hypotheses, and sections 7 and 8 test the
consequential hypotheses. Section 9 is the Discussion.
Copula approaches have been used to great effect in neuroscience [@onken2009], 
bioinformatics [@Kim2008], medical research [@Emura2016], direct study of environmental 
variables [@Serinaldi2008; @She2018; @Goswami2018; @li2013], and finance [@Li2000]. 
To our knowledge, these concepts have never been used in ecology. 
We argue that benefits to ecology of using a complete 
copula description of dependence will be substantial.

@@

dt <- tibble(Description = c("1. p-value, independence test",
           "2. Minimum-AIC copula",
           "3. AIC weight of minimum-AIC copula",
           "4. AIC weight of N copula",
           "5. p-value, Cramer-von Mises goodness of fit test",
           "6. p-value, Kolmogorov-Smirnov goodness of fit test",
           "7. Model averaged lower-tail dependence",
           "8. Model averaged upper-tail dependence",
           "9. Model averaged lower- minus upper-tail dependence",
           "10. Asymmetry rank in surrogates")
)
tab_bivar_summary<-cbind(dt,tab_bivar_summary)
knitr::kable(tab_bivar_summary[1:9,], 
             format="latex",
             caption = "Summary of results of testing the observational hypothesis for bivariate datasets. The $p$-values in rows 5-6 are for the minimum-AIC copula. Model averaging used for rows 7-9 was based on AIC weights. Cedar Creek results are for the year 2000.\\label{tab_bivar_summary}",
             booktabs=T, linesep = "\\addlinespace",
             col.names = c(" ", "Soil C and N", "Bird masses and BMR", "Mammal masses and BMR", "Cedar Creek data")
             )%>%column_spec(2:5,width="1.4cm")%>%
             column_spec(1, width = "8.5cm")%>%kable_styling(latex_options = "basic",position = "center")
```

@@

For the model we examined, causal hypothesis D (Fig. \ref{MasterFigure}) was 
correct that asymmetric tail dependence in evolutionary changes 
can produce similarly asymmetric tail dependence in character distributions across
phylogeny tips (Appendix \ref{SM-evol_mech} fro details). This topic is revisited in the Discussion.

@@

<!--what we did and why is it important to study?-->
We showed that non-N copula structure and asymmetric tail dependence are 
common in datasets across multiple sub-disciplines
and levels of organization in ecology (Fig. \ref{MasterFigure}, O), although no published 
research we are aware of has 
taken this structure into account. We hypothesized four 
mechanisms (Fig. \ref{MasterFigure}, A-D) that may cause non-N copula structures. One
mechanism (Fig. \ref{MasterFigure}, A) was via nonlinear environmental influences
on ecological variables. A second mechanism (Fig. \ref{MasterFigure}, B) produces tail dependence via
a Moran effect, whereby dependencies between environmental variables 
beget similar dependencies between ecological variables in different locations. 
We discuss below how commonly these and other
potential mechanisms may opperate. 
We also hypothesized and demonstrated important consequences of 
non-N and tail-dependent copula structure in ecology 
(Fig. \ref{MasterFigure}, X-Z). First (Fig. \ref{MasterFigure}, X), the skewness
of a spatial-average time series is influenced by tail dependencies between its 
constituent time series. Thus right-tail-dependent local time series 
lead to "spiky" spatially averaged time series, with large outbreaks.
These observations could have implications for numerous exploited and pest species
for which the spatial average time series have applied importance.
Second (Fig. \ref{MasterFigure}, Y), metapopulation extinction risk can be accentuated or reduced by tail dependence
properties of environmental variables. We also showed effects of copula 
structure on Taylor's law (Fig. \ref{MasterFigure}, Z), which has been applied in many fields.
In our view, our results make our "overall hypothesis," framed in the Introduction,
reasonable and worthy of broader testing by the research community. We have illustrated 
extensively how such tests might be carried out by performing them ourselves on 
example datasets and models. Further suggestions for future possible tests are below. 
<!--***DAN: make sure you have enough of this, Dan-->
It already seems likely that comprehensive understanding 
of many ecological phenomena is impossible without 
a complete copula characterization of dependence between variables.

@@

<!--a para on how likely one is to get copula structure via the Liebig mechanism-->
Our first causal mechanism (Fig. \ref{MasterFigure}, A) seems likely to operate commonly.
Liebig's law and the idea of a limiting 
nutrient of growth are dominant paradigms in ecology.
Although we did not systematically test the first two parts of causal 
hypothesis A, many studies document nonlinear influences of environmental
variables on ecological quantities at all levels of organization. 
These parts of A could be tested
straightforwardly for numerous datasets by examing copulas describing dependencies
between environmental and ecological
variables (see below discussion about the green spruce aphid, and 
Appendix \ref{SM-sect:greenspruce}, which illstrate how to do this).
Causal mechanism C, which we also did not test, could be tested straightforwardly
by analyzing copulas of abundances of competing species, across space or
through time.

<!--para on how likely one is to get copula structure via the Moran mechanism-->
There are also reasons to expect that our Moran mechanism (Fig. \ref{MasterFigure}, B) for 
copula structure and tail dependence will
operate commonly, because Moran effects are common [@Liebhold2004], and 
non-N and tail-dependent copula structures are often found in 
environmental variables. 
If large meteorological events are 
also spatially widespread, then
one would expect environmental variables associated with these events to take extreme
values simultaneously across large spatial areas, producing tail dependence in measurements
made through time at different locations. Non-extreme values, not associated with 
spatially extensive weather events, may
be more associated with local phenomena, and therefore 
less prone to being correlated across large areas.
@Serinaldi2008 examined the spatial dependence 
of rainfall observations in Central Italy; G or Student 2-copulas 
were possible candidates for modelling dependence, and 
neither of these is an N copula.
A long-term study (1950–2014) in the Loess Plateau of China [@She2018]
showed that a G copula (which has upper tail dependence) could effectively model 
the spatial dependence of drought variables in the study area.
Bivariate copula analysis was also used in forecasting the 
co-occurrence of extreme events (flood or drought) 
over the North Sikkim Himalayas, India, using spatial datasets [@Goswami2018].

<!--Holder stuff-->
Our simulations of character evolution suggest the hypothesis that evolutionary
changes through time in bird and mammal BMR and body size may exhibit right-tail dependence,
contrary to standard assumptions of models of character evolution. This is
a hypothesis only, because the right-tail dependence of Fig. \ref{fig_biv_multi_raw_cop}F,G
could have come about in another, unknown way: our simulations show that tail dependence 
in evolutionary changes is sufficient, but may not be necessary, to produce tail
dependence in characters of extant species. For instance, as discussed below and in
Appendix \ref{SM-sect:missingdata} (also Fig. \ref{SM-fig:missingdata}), 
systematically missing data can produce tail dependence and may
have influenced results for the BMR-body mass datasets. Even if the hypothesized 
evolutionary mechanism (Fig. \ref{MasterFigure}, D)
is correct, our results simply replace one question, i.e., why do we see right-tail
dependence between BMR and body mass for birds and mammals, with another question,
i.e., why might we see right-tail dependence in evolutionary changes in these traits?
See Appendix \ref{SM-evol_mech} for additional thoughts on the potential importance of 
copulas for studies of character evolution.

<!--Discuss stuff here on tradeoff and life history that Bever told me about-->
The relationships we studied between BMR and body mass relate
to a trade-off between mass-specific BMR (i.e., BMR per unit body mass) and body
mass itself. Trade-offs between life history characteristics are a focal topic in 
ecology. Copulas probably interrelate with 
life-history theory in additional ways beyond what we could demonstrate 
with the data we used.
It is well known that energy allocation to a life function, X (e.g., reproduction)
will reduce the energy that can be allocated to other functions, $Y_1$, $Y_2$, $Y_3$ 
(e.g., growth, predation avoidance; this is called the principle of 
allocation).
But it is equally well known that X can trade off against any or all of 
the $Y_i$. Therefore, for large values of X which approach absolute 
limitations, there may be a strong 
association between X and $Y_1$. Whereas for small values of X there may be
little association - resources not allocated to X can instead be allocated 
to $Y_1$, $Y_2$, $Y_3$ or to any combination thereof. This constitutes 
asymmetric tail dependence between X and $Y_1$. For instance, @Winemiller1992
described a three-way trade-off in North American fishes between age of reproductive
maturity, juvenile survivorship, and fecundity. The trade-off should, in theory,
produce a tight association between age of maturity and fecundity for 
fishes with low age of maturity, but little association between those
variables for later-maturing
fishes because such fishes may invest much or little in juvenile survival.
Although copulas may be very useful for studying multi-dimensional trade-offs, 
applications will 
require careful attention to the possible consequences of biased sampling:
if the degree of completeness of a dataset is associated with one or
more of the characters being studied then statistical artefacts
may bias conclusions (Appendix \ref{SM-sect:missingdata}, 
Fig. \ref{SM-fig:missingdata}).

@@

<!--green spruce aphid all the way from causes to consequences-->
Our hypotheses cover propagation of copula structure from causes, 
to observable data, to consequences for human concerns.
We here take a closer look at the green spruce aphid, which illustrates 
all these stages in one example.
Green spruce aphid abundance, as detected in the suction trap data we use (Data,
Appendix \ref{SM-Data_details}),
is strongly influenced by temperature the previous winter (@Sheppard2016,
supplementary fig. 6).
For each of our 10 sampling locations, we therefore 
examined (Appendix \ref{SM-sect:greenspruce}, 
Fig. \ref{SM-fig:greenspruce}, Table \ref{SM-tab:greenspruce}) the copula of winter temperature
and green spruce aphid abundance time series for the location, finding 
substantial left-tail dependence: apparently winter temperature has an asymmetric
influence on abundance in the sense that cold winters produce low abundances
but warm winters do not yield especially higher abundances than moderate winters.
Although we did not systematically test the first two 
parts of causal hypothesis A due to much existing evidence, this result 
illustrates how parts one and two of the hypothesis can be tested using copula
techniques. The third part of causal hypothesis A, which our modelling results
supported, now suggests spatial dependence between green spruce aphid counts
should exhibit left-tail dependence - exactly what was observed 
(Table \ref{tab_multivar_summary}). And the consequences of that spatial tail 
dependence for skewness of spatially averaged aphid counts was described already 
(Fig. \ref{fig_skewness_spearman}A). Thus the asymmetric influence of winter 
temperature ultimately causes spatially averaged green spruce aphid abundances to have
lower skew than they would otherwise, because negative impacts of cold winter
are spatially synchronous but positive impacts of warmer winters and other factors
are less spatially or temporally consistent.

<!--***DAN: could remove the next para if you need to save words-->
<!--a para on stability of the total biomass time series in a community, and skewness - sets up Shya's later paper as future work-->
Our results showed that the skewness of the 
spatial average of local time series from different locations is influenced 
by their tail dependence. But the logic of this result should also apply to any collection of time
series, whether associated with different locations in space or not, that are summed or averaged. 
Another potential application is time series of all species from a single community, e.g., all
plants in a quadrat surveyed repeatedly over time. A huge 
literature has focused on synchrony versus 
compensatory dynamics between such time series, and 
the influence of interspecific relationships on the 
variability of total-community
properties such as total biomass. Typically, variability through time of total-community
properties is measured with the coefficient of variation, but the skewness of such a time series
may also be of interest because it can help characterize "spikiness". Future work on copula 
structure of interspecific relationships in communities and its implications for community 
stability seems likely to be an exciting avenue of research.
